Project data cleaning

### 2024-05-03 Rcode/dneda-code/dneda.qmd
improve after review

This Quarto file loads the cleaned data and does some exploring.

I'm only showing it the way where the code is included in the file. 
As described in the `processing_code` materials, I currently prefer the approach of having R code in a separate file and pulling it in.

But I already had this written and haven't yet re-done it that way. Feel free to redo and send a pull request on GitHub :)

Again, it is largely a matter of preference and what makes the most sense to decide if one wants to have code inside Quarto files, or as separate R files.
And sometimes, an R script with enough comments is good enough and one doesn't need a Quarto file.

Also note that while here I split cleaning and exploring, this is iterative. You saw that as part of the processing, we already had to explore the data somewhat to understand how to clean it. In general, as you explore, you'll find things that need cleaning. As you clean, you can explore more. Therefore, at times it might make more sense to combine the cleaning and exploring code parts into a single R or Quarto file. Or split things in any other logical way.

As part of the exploratory analysis, you should produce plots or tables or other summary quantities for the most interesting/important quantities in your data. Depending on the total number of variables in your dataset, explore all or some of the others. Figures produced here might be histograms or density plots, correlation plots, etc. Tables might summarize your data.

Start by exploring one variable at a time. Then continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in. If your dataset is small, you can do that for all variables.

Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones.

>>>>

This file loads the processed data cleaned from processingfile-v1 and basic descriptive analysis is applied in this file.

### 2024-05-03 Rcode/mlmodel-code/ml-model-v1.qmd
improve after review

This Quarto file loads the cleaned data and does some exploring.

I'm only showing it the way where the code is included in the file. As described in the `processing_code` materials, I currently prefer the approach of having R code in a separate file and pulling it in.

But I already had this written and haven't yet re-done it that way. Feel free to redo and send a pull request on GitHub :)

Again, it is largely a matter of preference and what makes the most sense to decide if one wants to have code inside Quarto files, or as separate R files. And sometimes, an R script with enough comments is good enough and one doesn't need a Quarto file.

Also note that while here I split cleaning and exploring, this is iterative. You saw that as part of the processing, we already had to explore the data somewhat to understand how to clean it. In general, as you explore, you'll find things that need cleaning. As you clean, you can explore more. Therefore, at times it might make more sense to combine the cleaning and exploring code parts into a single R or Quarto file. Or split things in any other logical way.

As part of the exploratory analysis, you should produce plots or tables or other summary quantities for the most interesting/important quantities in your data. Depending on the total number of variables in your dataset, explore all or some of the others. Figures produced here might be histograms or density plots, correlation plots, etc. Tables might summarize your data.

Start by exploring one variable at a time. Then continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in. If your dataset is small, you can do that for all variables.

Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones.

@@

```{r}
# Split data into training and testing sets
set.seed(123)  # Set seed for reproducibility
trainIndex <- createDataPartition(mydata$BMI, p = 0.8, list = FALSE)
trainData <- mydata[trainIndex, ]
testData <- mydata[-trainIndex, ]
# Create indices for 10-fold cross-validation
cv_folds <- createFolds(trainData$BMI, k = 10)
```

@@

@@

@@

@@

```{r}
# Initialize variables to store results
min_rmse_lasso <- Inf
best_subset_lasso <- NULL

# Perform LASSO regression with cross-validation
lasso_model <- cv.glmnet(as.matrix(trainData[, predictors_updated]), trainData$BMI, alpha = 1, nfolds = 10)

@@

# Fit the LASSO model using the optimal lambda
lasso_model_optimal <- glmnet(as.matrix(trainData[, predictors_updated]), trainData$BMI, alpha = 1, lambda = optimal_lambda)

# Compute RMSE on test data using the optimal LASSO model
predictions_lasso <- predict(lasso_model_optimal, newx = as.matrix(testData[, predictors_updated]))
rmse_lasso <- sqrt(mean((testData$BMI - predictions_lasso)^2))

# Extract coefficients of the LASSO model
lasso_coefficients <- as.matrix(coef(lasso_model_optimal))

# Print results
cat("Optimal lambda:", optimal_lambda, "\n")
cat("RMSE with LASSO:", rmse_lasso, "\n")

# Print non-zero coefficients (selected predictors)
selected_predictors <- rownames(lasso_coefficients)[lasso_coefficients != 0, drop = FALSE]
print("Selected predictors by LASSO:")
print(selected_predictors)
```

After that, I will select another regression model to fit data, here I will use random forest model.

@@

  # Forward selection
  for (i in seq_along(all_predictors)) {
    predictors_subset_rf <- all_predictors[1:i]
    formula_subset_rf <- as.formula(paste(response, "~", paste(predictors_subset_rf, collapse = " + ")))

    # Fit random forest model using the training data for the current fold
    model_rf <- randomForest(formula = formula_subset_rf, data = mydata[train_indices, ])

    # Make predictions on the validation data for the current fold
    predictions_rf <- predict(model_rf, newdata = mydata[validation_indices, ])

    # Calculate RMSE for the current fold
    rmse_rf <- sqrt(mean((mydata$BMI[validation_indices] - predictions_rf)^2))

@@

      # Fit the DA model using the subset of predictors on the training data for the current fold
      formula_DA <- as.formula(paste(outcome, "~", paste(predictors_subset_DA, collapse = " + ")))
      model_DA <- lda(formula_DA, data = mydata[train_indices, ])

@@

      # Calculate accuracy for the current fold
      fold_accuracies[fold_index] <- mean(predicted_DA$class == mydata$Obesity[validation_indices])
    }

@@

+-----------------------------------+-------------------------------------------+---------------+------------------------------+
| Model                             | Predictor                                 | Metric Value  | Metric Value on testing data |
+===================================+===========================================+===============+==============================+
| Linear null                       | \-                                        | RMSE=8.00     | RMSE=8.04                    |
+-----------------------------------+-------------------------------------------+---------------+------------------------------+
| Linear regression 1               | Age, History, Alcohol, MTRANS             | RMSE=0.62     | RMSE=6.61                    |
+-----------------------------------+-------------------------------------------+---------------+------------------------------+
| Linear regression 2               | All                                       | RMSE=6.30     | RMSE=3.99                    |
+-----------------------------------+-------------------------------------------+---------------+------------------------------+
| Linear regression 2 + LASSO model | All                                       | RMSE=7.38     | RMSE=7.38                    |
|                                   |                                           |               |                              |
|                                   |                                           | l=0.0011      |                              |
+-----------------------------------+-------------------------------------------+---------------+------------------------------+
| Random forest                     | Gender, Age, History, Water, Alcohol, FAF | RMSE=3.22     | RMSE=3.97                    |
+-----------------------------------+-------------------------------------------+---------------+------------------------------+
| Logistic null                     | \-                                        | Accuracy=0.17 |                              |
+-----------------------------------+-------------------------------------------+---------------+------------------------------+
| Discriminant analysis             | All                                       | Accuracy=0.50 | Accuracy=0.48                |
+-----------------------------------+-------------------------------------------+---------------+------------------------------+
| Multinominal logistic regression  | All                                       | Accuracy=0.58 | Accuracy=0.52                |
+-----------------------------------+-------------------------------------------+---------------+------------------------------+

Then I will use testData to assess the quality to the model that I choose.

@@

@@

Here I try to assess every model I choose and check the result.

@@

# Fit the LASSO model using the optimal lambda
lasso_model_optimal <- glmnet(as.matrix(trainData[, predictors_updated]), trainData$BMI, alpha = 1, lambda = 0.0011)

# Make predictions on the test data using the optimal LASSO model
predictions_lasso_test <- predict(lasso_model_optimal, newx = as.matrix(testData[, predictors_updated]))

# Calculate RMSE on the test data
rmse_lasso_test <- sqrt(mean((testData$BMI - predictions_lasso_test)^2))

@@

# Define formula for DA model
formula_DA_test <- as.formula(paste("Obesity ~", paste(predictors_updated, collapse = " + ")))

@@

# Print the accuracy
print(paste("Accuracy of the DA model on test data:", accuracy_DA_test))
```

# Notes

For your own explorations, tables and figures can be "quick and dirty". As long as you can see what's going on, there is no need to polish them. That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.). Those should look as nice, polished and easy to understand as possible.

>>>>

@@

```{r}
# Filter out observations where Alcohol is 4
mydata1 <- mydata[mydata$Alcohol != 4, ]
# Split data into training and testing sets
set.seed(123)  # Set seed for reproducibility
trainIndex <- createDataPartition(mydata$BMI, p = 0.8, list = FALSE)
trainData <- mydata1[trainIndex, ]
testData <- mydata1[-trainIndex, ]
# Create indices for 10-fold cross-validation
cv_folds <- createFolds(trainData$BMI, k = 10)
```

Since there is only one observation with alcohol consumption level at 4, so after splitting the data, it will be distributed randomly in either trainData or testData. When it is in trainData and the cross-validation is applied, there will be some folds doesn't have this level and error will occur during this process. So I drop it from the dataset to make it working easier.

I will fit the first linear model which is called linear model 1 to the data and use cross-validation, then calculate RMSE. There is no interactions in this model's predictors. And by running this chunk, predictor combination that indicates the lowest RMSE will be selected and shown in the output together with its RMSE.

@@

# Set seed
set.seed(123)

@@

Then I will calculate RMSE for null model of linear regression.

@@

Using the same linear regression named linear model 2 but adding 2 interactions as predictors and under these predictors, best combination with lowest RMSE will be shown in the output.

@@

```{r}
# Create dummy variables for categorical variables in trainData
trainData1 <- within(trainData, {
  History <- as.factor(History)
  MTRANS <- as.factor(MTRANS)
})
dummy_vars_train <- model.matrix(~ History + MTRANS - 1, data = trainData1)

# Combine dummy variables with original trainData
trainData1 <- cbind(trainData, dummy_vars_train)

# Repeat the same process for testData
testData1 <- within(testData, {
  History <- as.factor(History)
  MTRANS <- as.factor(MTRANS)
})
dummy_vars_test <- model.matrix(~ History + MTRANS - 1, data = testData1)
testData1 <- cbind(testData, dummy_vars_test)

# Define predictors and outcome
response <- "BMI"
predictors <- c("Gender", "Age", "History", "Water", "Alcohol", "FAF", "MTRANS")
interaction_terms <- c("Gender_Age", "FAF_Water")

# Update predictors to include dummy variables
predictors_lasso <- c("Gender", "Age", "Water", "Alcohol", "FAF","Gender_Age", "FAF_Water", colnames(dummy_vars_train))

# Define function to calculate RMSE
calculate_rmse <- function(predictions, actual) {
  return(sqrt(mean((actual - predictions)^2)))
}

# Initialize variables to store results
min_rmse_lasso <- Inf
best_subset_lasso <- NULL

# Set seed
set.seed(123)

# Perform LASSO regression with cross-validation
for (fold in 1:10) {
  # Extract indices for training and validation data for current fold
  train_indices <- cv_folds[[fold]]
  validation_indices <- setdiff(1:nrow(trainData1), train_indices)

  # Define training and validation data subsets
  train_data_fold <- trainData1[train_indices, ]
  validation_data_fold <- trainData1[validation_indices, ]

  # Create design matrix for training and validation data
  train_matrix <- model.matrix(~., data = train_data_fold[, predictors_lasso])[,-1]
  validation_matrix <- model.matrix(~., data = validation_data_fold[, predictors_lasso])[,-1]

  # Ensure the same variables are present in training and validation matrices
  common_variables <- intersect(colnames(train_matrix), colnames(validation_matrix))
  train_matrix <- train_matrix[, common_variables]
  validation_matrix <- validation_matrix[, common_variables]

  # Perform LASSO regression with cross-validation
  lasso_model <- cv.glmnet(train_matrix, train_data_fold$BMI, alpha = 1, lambda = 10^seq(10, -2, length = 100))

  # Extract optimal lambda value
  optimal_lambda <- lasso_model$lambda.min

  # Fit LASSO model using optimal lambda on training data
  lasso_model_optimal <- glmnet(train_matrix, train_data_fold$BMI, alpha = 1, lambda = optimal_lambda)

  # Make predictions on validation set
  predictions_lasso <- predict(lasso_model_optimal, newx = validation_matrix, s = optimal_lambda)

  # Calculate RMSE on validation set
  rmse <- sqrt(mean((validation_data_fold$BMI - predictions_lasso)^2))

  # Update min_rmse and best_subset if CV RMSE is lower
  if (rmse < min_rmse_lasso) {
    min_rmse_lasso <- rmse
    best_subset_lasso <- predictors_lasso  # Include interaction terms
  }
}

# Extract non-zero coefficients from the optimal LASSO model
non_zero_coeffs <- coef(lasso_model_optimal)
non_zero_coeffs <- non_zero_coeffs[-1]  # Remove intercept

# Find the indices of non-zero coefficients
non_zero_indices <- which(non_zero_coeffs != 0)

# Extract the names of predictors with non-zero coefficients
best_subset_names_lasso <- colnames(train_matrix)[non_zero_indices]

@@

# Print results
cat("Lowest RMSE (LASSO CV):", min_rmse_lasso, "\n")
cat("Best subset of predictors:", paste(best_subset_names_lasso, collapse = ", "), "\n")
# Print optimal lambda value
cat("Optimal lambda:", optimal_lambda, "\n")
```

Since I already get optimal lambda above, so I will skip using autoplot function creating figure and tuning the LASSO model.

For trying with more models as comparison, I select another regression model to fit data, here I will use random forest model.

@@

  # Forward selection
  for (i in seq_along(predictors_updated)) {
    predictors_subset_rf <- predictors_updated[1:i]
    formula_subset_rf <- as.formula(paste(response, "~", paste(predictors_subset_rf, collapse = " + ")))

    # Fit random forest model using the training data for the current fold
    model_rf <- randomForest(formula = formula_subset_rf, data = trainData[train_indices, ])

    # Make predictions on the validation data for the current fold
    predictions_rf <- predict(model_rf, newdata = trainData[validation_indices, ])

    # Calculate RMSE for the current fold
    rmse_rf <- sqrt(mean((trainData$BMI[validation_indices] - predictions_rf)^2))

@@

      # Fit the DA model using the subset of predictors on the training data for the current fold
      formula_DA <- as.formula(paste(outcome, "~", paste(predictors_subset_DA, collapse = " + ")))
      model_DA <- lda(formula_DA, data = trainData[train_indices, ])

@@

      # Calculate accuracy for the current fold
      fold_accuracies[fold_index] <- mean(predicted_DA$class == trainData$Obesity[validation_indices])
    }

@@

+-----------------------------------+---------------------------------------------------------------+---------------+------------------------------+
| Model                             | Predictor                                                     | Metric Value  | Metric Value on testing data |
+===================================+===============================================================+===============+==============================+
| Linear null                       | \-                                                            | RMSE=8.01     | RMSE=7.99                    |
+-----------------------------------+---------------------------------------------------------------+---------------+------------------------------+
| Linear regression 1               | Age, History, Alcohol, MTRANS                                 | RMSE=0.62     | RMSE=6.54                    |
+-----------------------------------+---------------------------------------------------------------+---------------+------------------------------+
| Linear regression 2               | All                                                           | RMSE=6.29     | RMSE=3.81                    |
+-----------------------------------+---------------------------------------------------------------+---------------+------------------------------+
| Linear regression 2 + LASSO model | All                                                           | RMSE=6.34     | RMSE=6.47                    |
|                                   |                                                               |               |                              |
|                                   |                                                               | l=0.023       |                              |
+-----------------------------------+---------------------------------------------------------------+---------------+------------------------------+
| Random forest                     | All                                                           | RMSE=3.35     | RMSE=3.97                    |
+-----------------------------------+---------------------------------------------------------------+---------------+------------------------------+
| Logistic null                     | \-                                                            | Accuracy=0.17 | Accuracy=0.15                |
+-----------------------------------+---------------------------------------------------------------+---------------+------------------------------+
| Discriminant analysis             | Gender, Age, History, Water, Alcohol, FAF, MTRANS, Gender_Age | Accuracy=0.50 | Accuracy=0.50                |
+-----------------------------------+---------------------------------------------------------------+---------------+------------------------------+
| Multinominal logistic regression  | All                                                           | Accuracy=0.55 | Accuracy=0.50                |
+-----------------------------------+---------------------------------------------------------------+---------------+------------------------------+

Then I will use testData to assess the quality to the model that I choose. Between 2 linear regression models I choose random forest model.

@@

Between 2 logistic models, I choose mlr model.

@@

Since the neither of the model above has satisfying performance, I will try to assess every model I have used for analysis and check the result.

@@

# Fit the LASSO model using the optimal lambda
lasso_model_optimal <- glmnet(as.matrix(trainData1[, predictors_lasso]), trainData1$BMI, alpha = 1, lambda = 0.023)

# Make predictions on the test data using the optimal LASSO model
predictions_lasso_test <- predict(lasso_model_optimal, newx = as.matrix(testData1[, predictors_lasso]))

# Calculate RMSE on the test data
rmse_lasso_test <- sqrt(mean((testData1$BMI - predictions_lasso_test)^2))

@@

# Create new predictors for DA model
predictors_DA <- c("Gender", "Age", "History",  "Water",  "Alcohol", "FAF",  "MTRANS", "Gender_Age")

# Define formula for DA model
formula_DA_test <- as.formula(paste("Obesity ~ ", paste(predictors_DA, collapse = " + ")))

@@

# Print the accuracy
print(paste("Accuracy of the DA model on test data:", accuracy_DA_test))
```
I will explore more information for LASSO model.
```{r}
# Extract coefficients from the fitted LASSO model
lasso_coefficients <- coef(lasso_model_optimal)

# Extract non-zero coefficients and their corresponding predictors
non_zero_indices <- which(lasso_coefficients != 0, arr.ind = TRUE)
non_zero_coefficients <- lasso_coefficients[non_zero_indices]

# Extract names of predictors with non-zero coefficients
strongest_predictors <- rownames(lasso_coefficients)[non_zero_indices[, "col"]]

# Find the index of the strongest and weakest coefficients
strongest_index <- which.max(abs(non_zero_coefficients))
weakest_index <- which.min(abs(non_zero_coefficients))

# Extract the strongest and weakest coefficients and their predictors
strongest_coefficient <- non_zero_coefficients[strongest_index]
weakest_coefficient <- non_zero_coefficients[weakest_index]
strongest_predictor <- strongest_predictors[strongest_index]
weakest_predictor <- strongest_predictors[weakest_index]

# Print the results
cat("Strongest predictor:", strongest_predictor, "\n")
cat("Strongest coefficient:", strongest_coefficient, "\n")
cat("Weakest predictor:", weakest_predictor, "\n")
cat("Weakest coefficient:", weakest_coefficient, "\n")
```

### 2024-05-03 Rcode/mlmodel-code/ml-model-v2.qmd
improve after review

This Quarto file loads the cleaned data and does some exploring.

This is essentially the same as the other `exploratory_analysis` Quarto file, but now the code is not inside this file. Instead, it is pulled in from the R script `exploratorycode.R` using the code chunk labels.

>>>>

This Quarto file has the content of machine learning model analysis.

This is essentially the same as the other `ml-model-v1` Quarto file, but now the code is not inside this file. Instead, it is pulled in from the R script `ml-model.R` using the code chunk labels.

### 2024-05-03 Rcode/processing-code/processingfile-v1.qmd
improve after review

This contains the same code and comments/information as `processingcode.R`.

This just shows it as Quarto file, to give you an idea how to do it in a setup that combines code and text in a single file.

See the other Quarto file for my currently preferred approach of pulling code from the R script into the Quarto file.

@@

```{r}
codebook <- here::here("data","raw-data","codebook.xlsx")
read_excel(codebook)
```
Several ways of looking at the data

The meaning of each variable can be checked [online](https://www.kaggle.com/datasets/yasserh/titanic-dataset/data).

@@

First generating a new variable BMI using height and weight, as well as change some variable names and categorical levels.

@@

Finally, we save the clean data as RDS file. I suggest you save your processed and cleaned data as RDS or RDA/Rdata files. This preserves coding like factors, characters, numeric, etc. If you save as CSV, that information would get lost. However, CSV is better for sharing with others since it's plain text. If you do CSV, you might want to write down somewhere what each variable is.

See here for some suggestions on how to store your processed data: http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata

@@

# Notes

Removing anyone who had "faulty" or missing data is one approach. It's often not the best. based on your question and your analysis approach, you might want to do cleaning differently (e.g. keep individuals with some missing information).

>>>>

@@

```{r}
codebook <- here::here("data","raw-data","codebook.xlsx")
read_excel(codebook)
```

Several ways of looking at the data

@@

First generating a new variable BMI using height and weight, as well as change some variable names and categorical levels. For more intuitive observation and writing, I will change gender categories to 1 and 2 and name alcohol consumption levels to from 1 to 4.

@@

@@
